---
title: "Introduction to rddsga"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rddsga}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`rddsga` implements weighted subgroup analysis for regression discontinuity (RD) designs. The key idea is that when subgroups differ in observed characteristics near the RD cutoff, a raw comparison of subgroup-specific RD estimates conflates the causal effect of the subgroup with the effect of correlated moderators. Inverse probability weighting (IPW) rebalances observed moderators across subgroups, isolating the subgroup-attributable component of the treatment effect difference.

## Installation

```r
# Install from GitHub
devtools::install_github("acarril/rddsga")
```

## Synthetic example

We use the bundled `rddsga_synth` dataset. The running variable `x` is centered
at zero; the cutoff is at 0. The binary subgroup indicator is `sgroup`. The
continuous moderator `m` has different means across subgroups near the cutoff.
True per-subgroup RD effects are 2 (G=0) and 4 (G=1).

```{r load}
library(rddsga)
data(rddsga_synth)
str(rddsga_synth)
```

### Step 1: Unweighted subgroup analysis

```{r noipsw, cache = TRUE}
fit_unw <- rddsga(
  y ~ 1 | sgroup,          # outcome ~ covariates | subgroup
  data      = rddsga_synth,
  running   = ~ x,
  bwidth    = 0.5,
  noipsw    = TRUE,         # no IPW — just separate local linear regressions
  bootstrap = FALSE
)
print(fit_unw)
```

### Step 2: IPW-weighted subgroup analysis

Include the moderator `m` so the propensity score model can balance it.

```{r ipsw, cache = TRUE}
fit_ipsw <- rddsga(
  y ~ m | sgroup,
  data      = rddsga_synth,
  running   = ~ x,
  bwidth    = 0.5,
  bootstrap = FALSE
)
summary(fit_ipsw)
```

The balance table shows that IPW eliminates the standardized difference in `m`
across subgroups at the cutoff.

### Step 3: Adding bootstrap inference

```{r bootstrap, cache = TRUE}
fit_boot <- rddsga(
  y ~ m | sgroup,
  data      = rddsga_synth,
  running   = ~ x,
  bwidth    = 0.5,
  bsreps    = 200,
  seed      = 42
)
print(fit_boot)
```

### Step 4: Fuzzy RD via 2SLS

The dataset includes a fuzzy treatment `D` with ~80% compliance above the
cutoff. Use `model = "iv"` to obtain 2SLS estimates.

```{r fuzzy, cache = TRUE}
fit_iv <- rddsga(
  y ~ m | sgroup,
  data      = rddsga_synth,
  running   = ~ x,
  bwidth    = 0.5,
  model     = "iv",
  fuzzy     = ~ D,
  bootstrap = FALSE
)
print(fit_iv)
```

## Accessing results

All estimates, standard errors, and confidence intervals are stored in the
returned object:

```{r access}
# Point estimates
coef(fit_boot)

# Variance-covariance matrix for (G=0, G=1)
vcov(fit_boot)

# 95% confidence intervals
confint(fit_boot)

# Number of observations within bandwidth
nobs(fit_boot)

# Bootstrap draws (B × 2 matrix)
head(fit_boot$bootstrap$draws)
```

## Key options

| Argument | Description |
|---|---|
| `kernel` | `"uniform"` (default), `"triangular"`, `"epanechnikov"` |
| `p` | Polynomial order (default `1`; `2` for local quadratic) |
| `m` | IPW mode: `2` = balance both groups (default), `1` = ATT G=1, `0` = ATT G=0 |
| `comsup` | Restrict to common propensity score support |
| `rbalance` | Balance table means: `0` = at cutoff (default), `1` = in sample |
| `block_var` | Column name for stratified block bootstrap |
| `vce` | Sandwich SE type: `"HC1"` (default), `"HC0"`, `"HC2"`, `"HC3"`, `"cluster"` |
| `ps_model` | `"logit"` (default) or `"probit"` for propensity score |
