---
title: "Introduction to wsga: difference-in-differences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to wsga (DiD)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`wsga` implements weighted subgroup analysis (the "Weighted Subgroup Analysis"
framework) for designs that need control variables for identification. This
vignette covers the sharp 2-period difference-in-differences (DiD) path. The
goal is to recover per-subgroup treatment effects when subgroups differ in
observed moderators: a raw comparison conflates the causal effect of the
subgroup with the effect of correlated moderators, and inverse probability
weighting (IPW) on subgroup membership corrects for that imbalance.

## Loading the data

We use the bundled `wsga_did_synth` dataset: a balanced two-period panel of
500 units (1,000 rows). The binary subgroup `sgroup`, the moderator `m`, and
the treatment indicator `D` are all constant within unit. `D` is independent of
`sgroup`, but `m` has a higher mean in `sgroup = 1` — exactly the imbalance
that IPW is designed to address. True per-subgroup effects are
`tau_0 = 1`, `tau_1 = 3` (true difference = 2).

```{r load}
library(wsga)
data(wsga_did_synth)
str(wsga_did_synth)
table(wsga_did_synth$sgroup, wsga_did_synth$time)
```

### Step 1: Unweighted DiD subgroup analysis

The unweighted estimator fits a long-form TWFE regression (unit and time
fixed effects) interacted with subgroup. With `noipsw = TRUE`, no IPW
correction is applied — coefficients reflect both the subgroup effect and
the contribution of imbalanced moderators.

```{r noipsw}
fit_unw <- wsga(
  y ~ 1 | sgroup,
  data       = wsga_did_synth,
  design     = "did",
  unit       = "unit",
  time       = "time",
  treat      = "D",
  noipsw     = TRUE,
  bootstrap  = FALSE
)
print(fit_unw)
```

Point estimates should be close to (`1`, `3`) with a difference near `2`.

### Step 2: IPW-weighted DiD subgroup analysis

Adding `m` on the left of `|` makes it a covariate in the outcome regression
**and** a balance variable for the propensity score model on `sgroup`. The
balance tables (printed by `summary()`) show the imbalance before and after
weighting.

```{r ipsw}
fit_ipsw <- wsga(
  y ~ m | sgroup,
  data       = wsga_did_synth,
  design     = "did",
  unit       = "unit",
  time       = "time",
  treat      = "D",
  bootstrap  = FALSE
)
summary(fit_ipsw)
```

The unweighted aggregate balance shows a large standardized difference in `m`
across subgroups; IPW reduces it to near zero. For DiD, the package also
reports a treated-only balance table (units with `D = 1`), since the
treatment effect is identified off the treated cells.

### Step 3: Cluster bootstrap inference

When `design = "did"`, `cluster_var` defaults to the `unit` variable, so the
bootstrap is a pairs-cluster bootstrap over units (Cameron–Gelbach–Miller).

```{r bootstrap, cache = TRUE}
fit_boot <- wsga(
  y ~ m | sgroup,
  data       = wsga_did_synth,
  design     = "did",
  unit       = "unit",
  time       = "time",
  treat      = "D",
  bsreps     = 100,
  seed       = 1
)
print(fit_boot)
```

The reported standard errors and CIs come from the empirical bootstrap
distribution; the message at the bottom of the print confirms how many
clusters were resampled.

## Key options (DiD)

| Argument      | Description |
|---|---|
| `design`      | Must be `"did"` to use the difference-in-differences path. |
| `unit`        | Column name of the unit identifier (panel ID). |
| `time`        | Column name of the time variable; must have exactly 2 unique values. |
| `treat`       | Column name of the binary treatment indicator (constant within unit). |
| `post_value`  | Value of `time` denoting the post period. Defaults to `max(time)`. |
| `cluster_var` | Column defining clusters for the bootstrap and analytical SE. Defaults to `unit` for DiD. |
| `inference`   | `"empirical"` (default with bootstrap), `"normal"`, or `"analytical"` (cluster-robust sandwich when `cluster_var` is set). |
