---
title: "Weighted Subgroup Analysis"
subtitle: "Monte Carlo Simulation Evidence -- RDD Path"
date: today
format:
  html:
    theme: flatly
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  execute-dir: file
---

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Results directory (relative to this file)
RES <- "simulation_results"

# Shared aesthetics
PAL <- c(Naive = "#c0392b", WSGA = "#2980b9", `Full saturation` = "#27ae60")
LTY <- c(Naive = "dashed",  WSGA = "solid",   `Full saturation` = "dotted")

theme_sim <- function(base = 13) {
  theme_minimal(base_size = base) +
  theme(
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(colour = "grey92"),
    strip.background  = element_rect(fill = "#f0f0f0", colour = NA),
    strip.text        = element_text(face = "bold", size = rel(0.95)),
    legend.position   = "bottom",
    legend.title      = element_blank(),
    plot.title        = element_text(face = "bold"),
    plot.subtitle     = element_text(colour = "grey40", size = rel(0.9)),
    axis.title        = element_text(size = rel(0.9)),
    plot.caption      = element_text(colour = "grey50", size = rel(0.78),
                                     hjust = 0, margin = margin(t = 8))
  )
}

# Beta_M labeller
bm_lab <- function(x) paste0("beta_M = ", x)
n_lab  <- function(x) paste0("n = ", x)
```

## Overview

This document reports four Monte Carlo scenarios that validate the core
statistical properties of `wsga` for **sharp regression discontinuity (RD)**
designs.  All scenarios use the same underlying data-generating process (DGP)
and are run with $R = 1{,}000$ replications per cell.  The script that
produced these results lives at `inst/run_simulations.R`.

## Data-generating process {#sec-dgp}

The DGP is a sharp RD with one observable moderator $M$ that is
imbalanced across subgroups:

$$
y_i = \underbrace{\bigl(2 + \beta_M M_i\bigr) Z_i (1 - G_i)}_{\text{G=0 treatment response}}
    + \underbrace{\bigl(2 + \beta_G + \beta_M M_i\bigr) Z_i G_i}_{\text{G=1 treatment response}}
    + 0.3\, x_i (1-G_i) + 0.5\, x_i G_i + \varepsilon_i
$$

| Term | Distribution | Role |
|------|------|------|
| $x_i$  | $U(-1,1)$   | Running variable |
| $Z_i = \mathbf{1}[x_i \ge 0]$ | Sharp | Treatment assignment |
| $G_i$  | $\text{Bern}(0.5)$ | Binary subgroup |
| $M_i = \delta G_i + \varepsilon_M$ | $\varepsilon_M \sim N(0,1)$ | Moderator (imbalanced by $\delta$) |
| $\varepsilon_i$ | $N(0, 0.25)$ | Idiosyncratic noise |

: {tbl-colwidths="[38,28,34]"}

**True per-group RD effects**: $b_{G=0} = 2$, $b_{G=1} = 2 + \beta_G$.
**True differential**: $\Delta = \beta_G = 2$ (held fixed throughout).

The key feature is that $M$ amplifies the *treatment response* (it multiplies
$Z$, not just $y$).  This means the within-group RD estimate for $G=1$ picks
up $\beta_M \cdot E[M \mid G=1] = \beta_M \delta$, creating a bias of
$\beta_M \delta$ in the naive cross-subgroup comparison.  IPW (WSGA)
removes this bias by reweighting $M$ to the same distribution across groups.

Two estimators are compared throughout:

- **Naive** -- `wsga_rdd(y ~ 1 | G, noipsw = TRUE)`: no adjustment for $M$.
- **WSGA** -- `wsga_rdd(y ~ 1 | G, balance = ~ M, noipsw = FALSE)`: $M$ enters
  the propensity score only; outcome regression remains lean.

Both use analytical inference (no bootstrap) for speed, except Scenario 1b
which uses the pairs bootstrap to assess coverage.

---

## Scenario 1 -- Bias removal {#sec-bias}

**Design**: sweep $\delta \in \{0, 0.5, 1, 2\}$ and
$\beta_M \in \{0, 0.3, 0.5\}$ at $n \in \{300, 600, 1500\}$.
Report **bias** (mean estimate $- \Delta$) and **RMSE**.

**Prediction**: naive bias $\approx \beta_M \delta$; WSGA bias $\approx 0$.

```{r s1-load}
s1 <- read.csv(file.path(RES, "s1_bias_rmse.csv")) |>
  mutate(
    beta_M_lab = factor(paste0("beta[M] == ", beta_M),
                        levels = paste0("beta[M] == ", sort(unique(beta_M)))),
    n_lab = factor(paste0("n = ", n), levels = paste0("n = ", sort(unique(n))))
  )
```

### Bias {.unnumbered}

```{r s1-bias-plot, fig.height=5}
s1_long <- s1 |>
  select(n, delta, beta_M, beta_M_lab, n_lab, bias_naive, bias_wsga) |>
  pivot_longer(c(bias_naive, bias_wsga),
               names_to = "method", values_to = "bias") |>
  mutate(method = recode(method, bias_naive = "Naive", bias_wsga = "WSGA"))

# Theoretical naive bias line
theory <- s1 |>
  select(n, delta, beta_M, beta_M_lab, n_lab) |>
  distinct() |>
  mutate(theory = beta_M * delta)

ggplot(s1_long |> filter(n == 600),
       aes(x = delta, y = bias, colour = method, linetype = method)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_line(data = theory |> filter(n == 600),
            aes(x = delta, y = theory, group = beta_M_lab),
            colour = "grey60", linetype = "dotted", linewidth = 0.8,
            inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ beta_M_lab, labeller = label_parsed, nrow = 1) +
  scale_colour_manual(values = PAL) +
  scale_linetype_manual(values = LTY) +
  scale_x_continuous(breaks = c(0, 0.5, 1, 2)) +
  labs(
    title    = "Scenario 1: Bias in the differential effect estimate",
    subtitle = "n = 600  |  1000 replications  |  dotted line = theoretical naive bias",
    x        = expression(delta ~ "(moderator imbalance)"),
    y        = "Bias (mean estimate - true)",
    caption  = "WSGA residual bias at delta = 2 reflects IPW weight variance under extreme imbalance."
  ) +
  theme_sim()
```

The naive estimator's bias grows almost exactly linearly with
$\beta_M \cdot \delta$ (confirmed by the dotted theoretical line).  WSGA
removes nearly all of it; the small residual at $\delta = 2$ reflects IPW
weight variance when propensity scores are extreme.

### RMSE {.unnumbered}

```{r s1-rmse-plot}
s1_rmse <- s1 |>
  select(n, delta, beta_M, beta_M_lab, n_lab, rmse_naive, rmse_wsga) |>
  pivot_longer(c(rmse_naive, rmse_wsga),
               names_to = "method", values_to = "rmse") |>
  mutate(method = recode(method, rmse_naive = "Naive", rmse_wsga = "WSGA"))

ggplot(s1_rmse |> filter(n == 600),
       aes(x = delta, y = rmse, colour = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ beta_M_lab, labeller = label_parsed, nrow = 1) +
  scale_colour_manual(values = PAL) +
  scale_linetype_manual(values = LTY) +
  scale_x_continuous(breaks = c(0, 0.5, 1, 2)) +
  labs(
    title    = "Scenario 1: Root mean squared error",
    subtitle = "n = 600  |  1000 replications",
    x        = expression(delta ~ "(moderator imbalance)"),
    y        = "RMSE"
  ) +
  theme_sim()
```

When $\beta_M = 0$ WSGA's RMSE slightly exceeds naive (IPW weight variance with
no bias to correct).  Once confounding is present ($\beta_M > 0$) WSGA
dominates -- it trades a small efficiency cost for large bias reduction.

---

## Scenario 1b -- Bootstrap coverage {#sec-coverage}

**Design**: 500 reps of WSGA with pairs bootstrap ($B = 100$) at
$\delta \in \{0, 1\}$, $\beta_M = 0.3$, $n \in \{300, 600, 1500\}$.
Target: 95% CI empirical coverage.

```{r s1b-plot, fig.height=3.8}
s1b <- read.csv(file.path(RES, "s1_coverage.csv")) |>
  mutate(delta_lab = paste0("delta = ", delta))

ggplot(s1b, aes(x = factor(n), y = coverage, colour = delta_lab,
                group = delta_lab)) +
  geom_hline(yintercept = 0.95, colour = "grey50", linetype = "dashed",
             linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 3.5) +
  annotate("text", x = 0.6, y = 0.952, label = "Nominal 0.95",
           colour = "grey40", size = 3.5, hjust = 0) +
  scale_y_continuous(limits = c(0.89, 0.97), labels = percent_format(1)) +
  scale_colour_manual(values = c("#16a085", "#8e44ad")) +
  labs(
    title    = "Scenario 1b: Empirical 95% CI coverage (WSGA, pairs bootstrap)",
    subtitle = "500 reps  |  B = 100 bootstrap replications  |  beta_M = 0.3",
    x        = "Sample size n",
    y        = "Empirical coverage",
    caption  = "Slight under-coverage is expected at B = 100; tightens with larger n."
  ) +
  theme_sim() +
  theme(legend.title = element_text())
```

Coverage is 0.918-0.944 across all cells -- modest under-coverage typical of
bootstrap CIs at $B = 100$.  Coverage is not sensitive to the imbalance
level $\delta$, suggesting the bias is adequately removed before the
bootstrap runs.

---

## Scenario 2 -- Size control {#sec-size}

**Design**: true differential $\Delta = 0$ (null); sweep
$\delta \in \{0, 0.5, 1, 2\}$ and $\beta_M \in \{0, 0.3\}$ at $n = 600$.
Report the 5%-level rejection rate.

**Prediction**: naive over-rejects when $\delta > 0$ and $\beta_M > 0$;
WSGA maintains nominal size.

```{r s2-plot, fig.height=4.5}
s2 <- read.csv(file.path(RES, "s2_size.csv")) |>
  mutate(beta_M_lab = factor(paste0("beta[M] == ", beta_M),
                              levels = paste0("beta[M] == ", sort(unique(beta_M)))))

s2_long <- s2 |>
  pivot_longer(c(size_naive, size_wsga),
               names_to = "method", values_to = "size") |>
  mutate(method = recode(method, size_naive = "Naive", size_wsga = "WSGA"))

ggplot(s2_long |> filter(n == 600),
       aes(x = delta, y = size, colour = method, linetype = method)) +
  geom_hline(yintercept = 0.05, colour = "grey50", linetype = "dashed",
             linewidth = 0.6) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  annotate("text", x = 0.05, y = 0.053, label = "Nominal alpha = 0.05",
           colour = "grey40", size = 3.5, hjust = 0) +
  facet_wrap(~ beta_M_lab, labeller = label_parsed, nrow = 1) +
  scale_colour_manual(values = PAL) +
  scale_linetype_manual(values = LTY) +
  scale_x_continuous(breaks = c(0, 0.5, 1, 2)) +
  scale_y_continuous(labels = percent_format(1)) +
  labs(
    title    = "Scenario 2: Rejection rate under the null (Delta = 0)",
    subtitle = "n = 600  |  1000 replications  |  nominal alpha = 0.05",
    x        = expression(delta ~ "(moderator imbalance)"),
    y        = "Rejection rate",
    caption  = paste0(
      "WSGA mild inflation at (delta = 2, beta_M = 0) reflects IPW weight variance ",
      "when there is no confounding to correct."
    )
  ) +
  theme_sim()
```

The size distortion of the naive estimator is dramatic: at
$(\delta = 1,\, \beta_M = 0.3)$ naive rejects 21.7% of nulls at the 5%
level; at $(\delta = 2,\, \beta_M = 0.3)$ it reaches 65.8%.  WSGA holds
close to 5% throughout -- the notable exception is at $(\delta = 2,\, \beta_M = 0)$
where IPW reweighting is unnecessary but still adds weight variance,
inflating size to 12.9%.  This is a meaningful caveat: when there is no
confounding, WSGA is not the right tool.

---

## Scenario 3 -- Power as a function of K {#sec-power}

**Design**: $K \in \{0, 1, 2, 4, 6, 8\}$ continuous moderators
($M_k \sim N(0.4\, G, 1)$) with no direct effect on $y$.  True
$\Delta = 0.5$.  Three estimators:

| Estimator | Outcome formula | Propensity score |
|-----------|-----------------|------------------|
| Naive     | `y ~ 1 \| G`   | None (noipsw)    |
| WSGA      | `y ~ 1 \| G`   | $M_1 + \ldots + M_K$ |
| Full saturation | `y ~ M1 + ... + MK \| G` | None (noipsw) |

Full saturation adds $2K$ df to the outcome regression;
WSGA adds $K$ logit parameters to the propensity score but keeps the
outcome lean.

> **Note on the comparison.** The Calonico et al. "full saturation"
> approach cells on all $2^K$ moderator combinations, adding $2^K$
> parameters -- far more than the $2K$ in our linear-covariate proxy.
> With $K = 8$ and $n = 600$, $2^8 = 256$ cells average fewer than 2.5
> observations each, making the approach infeasible.  We show the
> linear-covariate proxy as the most optimistic comparator available in
> this framework.

```{r s3-plot, fig.height=5}
s3 <- read.csv(file.path(RES, "s3_power_vs_K.csv")) |>
  mutate(n_lab = factor(paste0("n = ", n), levels = paste0("n = ", sort(unique(n)))))

s3_long <- s3 |>
  pivot_longer(c(power_naive, power_wsga, power_full),
               names_to = "method", values_to = "power") |>
  mutate(method = recode(method,
    power_naive = "Naive",
    power_wsga  = "WSGA",
    power_full  = "Full saturation"))

ggplot(s3_long, aes(x = K, y = power, colour = method, linetype = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ n_lab, nrow = 1) +
  scale_colour_manual(values = PAL) +
  scale_linetype_manual(values = LTY) +
  scale_x_continuous(breaks = c(0, 1, 2, 4, 6, 8)) +
  scale_y_continuous(labels = percent_format(1), limits = c(0, 1)) +
  labs(
    title    = "Scenario 3: Power vs number of moderators K",
    subtitle = "1000 replications  |  Delta = 0.5  |  moderators have no direct effect on y",
    x        = "Number of moderators K",
    y        = "Rejection rate (true Delta = 0.5, alpha = 0.05)",
    caption  = paste0(
      "WSGA pays an IPW weight variance cost as K grows. ",
      "Full saturation (linear 2K df) stays near naive with this DGP because ",
      "moderators add no residual variance reduction. ",
      "Cell-splitting (2^K parameters) would collapse far faster."
    )
  ) +
  theme_sim()
```

With no direct $y$-effect from the moderators, full saturation's 2K df
overhead is small relative to sample size, so its power barely moves.
WSGA's power declines more noticeably because the propensity score model
with K continuous moderators creates IPW weight variance even when the
weights are not needed to remove bias.

The honest take-away from S3 is that **WSGA's power advantage over full
saturation materialises when the Calonico et al. cell-splitting is the
correct comparator** -- where each additional moderator multiplies the
required cell count by two.  The linear-covariate proxy shown here is
the most favourable comparator, and even then WSGA loses on raw power
when confounding is absent.  Its value rests on S1 and S2: unbiasedness
and size control under confounding, at a modest but non-zero power cost.

---

## Scenario 4 -- Sensitivity to unobservable confounding {#sec-unobs}

**Design**: unobservable $U = \gamma G + \varepsilon_U$ ($\varepsilon_U \sim N(0,1)$,
independent of $M$) amplifies the treatment response via $U \cdot Z$.
Sweep $\gamma \in \{0, 0.25, 0.5, 1, 2\}$ at $n \in \{600, 1500\}$.
WSGA controls for $M$ but cannot observe $U$.

**Prediction**: residual bias $\approx \gamma$ (the $G$-correlated component
of $U$ that IPW cannot reach).

```{r s4-plot, fig.height=4.5}
s4 <- read.csv(file.path(RES, "s4_unobservable.csv")) |>
  mutate(n_lab = factor(paste0("n = ", n), levels = paste0("n = ", sort(unique(n)))))

# Reference line: y = gamma (theoretical residual bias if WSGA perfectly removes M-bias)
ref <- data.frame(gamma = c(0, 2), bias = c(0, 2))

ggplot(s4, aes(x = gamma, y = bias_wsga, colour = n_lab, group = n_lab)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = "dashed",
              linewidth = 0.7) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  annotate("text", x = 1.62, y = 1.5, label = "bias = gamma (theoretical)",
           colour = "grey45", size = 3.5, angle = 42) +
  scale_colour_manual(values = c(`n = 600` = "#e67e22", `n = 1500` = "#16a085")) +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 1, 2)) +
  labs(
    title    = "Scenario 4: Residual bias from unobservable confounding",
    subtitle = "1000 replications  |  delta = 1, beta_M = 0.3",
    x        = expression(gamma ~ "(unobservable amplification strength)"),
    y        = "WSGA bias (mean estimate - true Delta)",
    caption  = paste0(
      "Small offset above the diagonal (~0.04 at n = 600, ~0.03 at n = 1500) ",
      "is residual from the observable M-confounding at delta = 1."
    )
  ) +
  theme_sim() +
  theme(legend.title = element_text(size = rel(0.9)))
```

The bias grows almost perfectly linearly with $\gamma$.  The small offset
above the $\text{bias} = \gamma$ diagonal is the residual from the
observable confounding ($\delta = 1,\, \beta_M = 0.3$) that WSGA does
not fully remove at finite $n$.  This residual shrinks with $n$,
consistent with IPW consistency.  The practical implication: Assumption 4
(no unobservable treatment-response confounders correlated with $G$) is
load-bearing, and the simulation quantifies exactly how fast the bias
accumulates when it fails.

---

## Summary {#sec-summary}

```{r summary-table}
knitr::kable(
  data.frame(
    Scenario = c("S1 Bias removal", "S1b Bootstrap coverage",
                 "S2 Size control", "S3 Power vs K", "S4 Unobservables"),
    `Key result` = c(
      "Naive bias = beta_M x delta; WSGA near-zero",
      "0.92-0.94 (nominal 0.95); stable across delta",
      "Naive rejects 22%-66% under the null; WSGA holds at ~5%",
      "Naive power flat; WSGA declines with K (IPW cost)",
      "Residual bias grows linearly with gamma"
    ),
    `Limitation / caveat` = c(
      "Small residual at delta = 2 from IPW weight variance",
      "Mild under-coverage at B = 100; improves with larger B",
      "WSGA size inflated at (delta = 2, beta_M = 0) when IPW is unnecessary",
      "Full saturation (linear 2K df) stays near naive in this DGP",
      "Assumption 4 is load-bearing; no remedy within the model"
    ),
    check.names = FALSE
  ),
  col.names = c("Scenario", "Key result", "Limitation / caveat")
)
```

::: {.callout-tip}
## When to use WSGA

Use WSGA when:

1. Observable moderators predict subgroup membership ($\delta > 0$).
2. Those same moderators amplify the treatment response ($\beta_M > 0$).
3. Omitting them would distort the cross-subgroup comparison.

When neither condition holds, WSGA pays an IPW variance cost for no
bias benefit.  The simulation provides a rough guide: size distortion
becomes serious around $\delta \ge 0.5$ with $\beta_M = 0.3$; at
$\delta = 1,\, \beta_M = 0.5$ the naive estimator has RMSE twice that of
WSGA at $n = 600$.
:::

::: {.callout-note}
## Reproducibility

```r
# From the package root:
Rscript inst/run_simulations.R           # full run (~12 min, 9 cores)
Rscript inst/run_simulations.R --quick  # smoke run (~30 s)
quarto render simulation/simulation_report.qmd
```
:::
